### 2024-04-02 paper/paper.qmd
update paper/paper.qmd

---
title: "Understanding Suicide Rates: Exploring Gender and Age Patterns That Lead to Various Risk Factors Associated with Suicidal Behaviour"
subtitle: "My subtitle if needed"
author: 
  - Sarah Lee
thanks: "Code and data are available at: https://github.com/sarahhhh02/final-paper.git"
date: today
date-format: long
abstract: "This paper utilizes data from Statistics Canada, retrieving data related to the Suicide Rates. Analysing the relationship with gender, it is shown that suicide rates tend to be higher in male rather than female. However, suicide rates vary between gender through age, as we find that adolecent females may have a higher rate for suicidal attempts. The results of these findings aids policymakers, public health officials, and mental health professionals to develop a more effective and strategic suicide prevention tailored for the population."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
```

# Introduction

Suicide is a significant worldwide concern as it concerns individuals, friends, families, and communities. Understanding the factors of suicide is crucial as it extends beyond a loss of a life and onto ongoing world problems like mental health. Hence understanding factors that contribute to suicide is critical into developing prevention strategies within the public health officials and professionals.

The complexity of suicide roots from various factors such as vulnerabilities, social determinants, cultural norms, and mental health being the lead cause for suicides. Moreover, suicides are varied from different demographic groups such as age and gender. Which highlights the need for specified research that consider these variables with suicide. Analyzing the relationship between how these factors play part in suicide rates can help further come up with answers and measures for a better future.

In this paper, we aim to explore the issues of suicide, carefully focusing on how age and gender play into its rates through a linear regression. The estimand here is how age and gender are related to the risk factors of suicides. Meaning, how they play into determining which group of individuals are at high suicidal risk. We will draw data from Statistics Canada, specifically focusing in Canadian suicide rates in 2019.

# Data {#sec-data}

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it.

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```

\newpage

# References

### 2024-04-02 paper/paper.qmd
update paper/paper.qmd

@@
 title: "Understanding Suicide Rates: Exploring Gender and Age Patterns That Lead to Various Risk Factors Associated with Suicidal Behaviour"
-subtitle: "My subtitle if needed"
 author: 

### 2024-04-08 paper/paper.qmd
update paper/paper.qmd

@@
   - Sarah Lee
-thanks: "Code and data are available at: https://github.com/sarahhhh02/final-paper.git"
+thanks: "Code and data are available at: https://github.com/sarahhhh02/suicides.git"
 date: today
@@
 
+```{=tex}
 \begin{align} 
@@
 \end{align}
-
+```
 We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.
@@
 
-# Appendix {-}
+# Appendix {.unnumbered}
 

### 2024-04-09 scripts/01-download_data.R
update scripts/01-download_data.R

@@ -1,26 +1,20 @@
 #### Preamble ####
-# Purpose: Downloads and saves the data from [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Downloads and saves the data from https://data.cdc.gov
+# Author: Sarah Lee
+# Date: 8 April 2024
+# Contact: sarahhhh.lee@mail.utoronto.ca 
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
 
 
-#### Workspace setup ####
-library(opendatatoronto)
-library(tidyverse)
-# [...UPDATE THIS...]
+### Download Data ###
+# Proceed to your web browser and paste this URL "https://data.cdc.gov" 
+# Search in the CDC website "Death rates for suicide, by sex, race, Hispanic origin, and age: United State"
+# Click on the dataset with the name "Death rates for suicide, by sex, race, Hispanic origin, and age: United States". The dataset selected should be provided by NCHS.
+# Click on the "Export" button located on the top right of the page
+# Select the "Download file", then make sure to export data in CSV form and select "All data" if not selected
+# Click Download
 
-#### Download data ####
-# [...ADD CODE HERE TO DOWNLOAD...]
-
-
-
-#### Save data ####
-# [...UPDATE THIS...]
-# change the_raw_data to whatever name you assigned when you downloaded it.
-write_csv(the_raw_data, "inputs/data/raw_data.csv") 
+### Downloaded Data ###
+# Save donwloaded CSV into "data/raw_data/suicide_raw.csv"
 
          

### 2024-04-10 scripts/02-data_cleaning.R
update scripts/02-dat_cleaning.R

@@ -9,36 +9,33 @@
 
 #### Workspace setup ####
 library(tidyverse)
+library(dplyr)
+
 
 #### Clean data ####
-raw_data <- read_csv("inputs/data/plane_data.csv")
+raw_data <- read_csv("data/raw_data/Death_rates_for_suicide__by_sex__race__Hispanic_origin__and_age__United_States_20240410.csv")
 
-cleaned_data <-
-  raw_data |>
-  janitor::clean_names() |>
-  select(wing_width_mm, wing_length_mm, flying_time_sec_first_timer) |>
-  filter(wing_width_mm != "caw") |>
-  mutate(
-    flying_time_sec_first_timer = if_else(flying_time_sec_first_timer == "1,35",
-                                   "1.35",
-                                   flying_time_sec_first_timer)
-  ) |>
-  mutate(wing_width_mm = if_else(wing_width_mm == "490",
-                                 "49",
-                                 wing_width_mm)) |>
-  mutate(wing_width_mm = if_else(wing_width_mm == "6",
-                                 "60",
-                                 wing_width_mm)) |>
-  mutate(
-    wing_width_mm = as.numeric(wing_width_mm),
-    wing_length_mm = as.numeric(wing_length_mm),
-    flying_time_sec_first_timer = as.numeric(flying_time_sec_first_timer)
-  ) |>
-  rename(flying_time = flying_time_sec_first_timer,
-         width = wing_width_mm,
-         length = wing_length_mm
-         ) |> 
-  tidyr::drop_na()
+#Total suicide rate by sex
+cleaned_sex <- raw_data %>%
+  select(UNIT, STUB_NAME, STUB_LABEL, YEAR, AGE, ESTIMATE) %>%
+  filter(STUB_LABEL %in% c("Male", "Female"))
+
+#Total suicide rate by age range
+cleaned_age <- raw_data %>%
+  select(UNIT, STUB_NAME, STUB_LABEL, YEAR, AGE, ESTIMATE) %>%
+  filter(STUB_LABEL %in% c("10-14 years", "15-24 years", "20-24 years", "25-44 years", "35-44 years",
+                    "45-64 years", "55-64 years", "65 years and over", "65-74 years", "75-84 years", 
+                    "85 years and over"))
+
+#Total suicide rate by sex and race
+cleaned_sex_race <- raw_data %>%
+  select(UNIT, STUB_NAME, STUB_LABEL, YEAR, AGE, ESTIMATE) %>%
[…]

### 2024-04-10 scripts/01-download_data.R
update scripts/01-download_data.R

@@ -15,6 +15,6 @@
 # Click Download
 
 ### Downloaded Data ###
-# Save donwloaded CSV into "data/raw_data/suicide_raw.csv"
+# Save downloaded CSV into "data/raw_data/Death_rates_for_suicide__by_sex__race__Hispanic_origin__and_age__United_States_20240410.csv"
 
          

### 2024-04-11 scripts/00-simulate_data.R
update scripts/00-simulate_data.R

@@ -10,10 +10,25 @@
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
+library(dplyr)
 
 #### Simulate data ####
-# [...ADD CODE HERE...]
+# Set seed for reproducibility
+set.seed(123)
+
+# Create a data frame with Suicide Rates on Age, Sex and Race by Year of 1000 population
+data <- data.frame(
+  UNIT = rep("Deaths per 100,000 resident population, crude", 1000), # Assuming 1000 observations
+  YEAR = sample(1950:2018, 1000, replace = TRUE), # Random years between 1950 and 2018
+  SEX = sample(c("Male", "Female"), 1000, replace = TRUE), # Random sex
+  RACE = sample(c("White", "Black or African American", "Asian or Pacific Islander", "American Indian or Alaska Native"), 1000, replace = TRUE), # Random race
+  AGE = sample(c("10-14 years", "15-24 years", "20-24 years", "25-44 years","35-44 years", "45-64 years","55-64 years",
+                 "65 years and over", "65-74 years", "75-84 years", "85 years and over"), 1000, replace = TRUE), # Random age group
+  ESTIMATE = round(rnorm(1000, mean = 20, sd = 10), 1) # Random death rates with mean 20 and standard deviation 10 rounded to one decimal point
+)
+
+# See first few rows of data frame
+head(data)
 
 
 

### 2024-04-11 scripts/02-data_cleaning.R
update scripts/02-data_cleaning.R

@@ -1,41 +1,80 @@
 #### Preamble ####
-# Purpose: Cleans the raw plane data recorded by two observers..... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 6 April 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Cleans the raw suicides data 
+# Author: Sarah Lee
+# Date: 10 April 2024
+# Contact: sarahhhh.lee@mail.utoronto.ca 
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: Raw data on Suicide Rates
+
 
 #### Workspace setup ####
 library(tidyverse)
 library(dplyr)
+library(arrow)
 
 
 #### Clean data ####
 raw_data <- read_csv("data/raw_data/Death_rates_for_suicide__by_sex__race__Hispanic_origin__and_age__United_States_20240410.csv")
 
-#Total suicide rate by sex
+###Total suicide rate by sex###
 cleaned_sex <- raw_data %>%
-  select(UNIT, STUB_NAME, STUB_LABEL, YEAR, AGE, ESTIMATE) %>%
-  filter(STUB_LABEL %in% c("Male", "Female"))
+  select(UNIT, YEAR, STUB_LABEL, ESTIMATE) %>%
+  filter(STUB_LABEL %in% c("Male", "Female")) %>%
+  rename(SEX = STUB_LABEL)
 
-#Total suicide rate by age range
+write_parquet(cleaned_sex, "data/analysis_data/cleaned_sex.parquet")
+
+###Total suicide rate by age range###
 cleaned_age <- raw_data %>%
-  select(UNIT, STUB_NAME, STUB_LABEL, YEAR, AGE, ESTIMATE) %>%
+  select(UNIT, YEAR, STUB_LABEL, ESTIMATE) %>%
   filter(STUB_LABEL %in% c("10-14 years", "15-24 years", "20-24 years", "25-44 years", "35-44 years",
                     "45-64 years", "55-64 years", "65 years and over", "65-74 years", "75-84 years", 
-                    "85 years and over"))
+                    "85 years and over")) %>%
+  rename(AGE = STUB_LABEL) 
 
-#Total suicide rate by sex and race
[…]

### 2024-04-21 suicide_sex_age_race/app.R
update

@@ -1,0 +1,55 @@
+#
+# This is a Shiny web application. You can run the application by clicking
+# the 'Run App' button above.
+#
+# Find out more about building applications with Shiny here:
+#
+#    https://shiny.posit.co/
+#
+
+# Load required libraries
+library(shiny)
+library(ggplot2)
+
+
+# Define UI
+ui <- fluidPage(
+  titlePanel("Suicide Rates by Sex, Age and Race"),
+  sidebarLayout(
+    sidebarPanel(
+      selectInput("SEX", "Select Sex:", choices = c("Male", "Female")),
+      checkboxGroupInput("AGE", "Select Age Group:", 
+                         choices = c("15-24 years", "25-44 years", "45-64 years", "65 years and over", "75-84 years", "85 years and over"),
+                         selected = c("15-24 years", "25-44 years","45-64 years", "65 years and over", "75-84 years", "85 years and over")),
+      checkboxGroupInput("RACE", "Select Age Group:", 
+                         choices = c("White", "Black or African American", "American Indian or Alaska Native", "Asian or Pacific Islander"),
+                         selected = c("White", "Black or African American", "American Indian or Alaska Native", "Asian or Pacific Islander"))
+    ),
+    mainPanel(
+      plotOutput("plot")
+    )
+  )
+)
+
+server <- function(input, output) {
+  
+  # Retrieve data from file
+  cleaned_sex_age_race <- read_parquet(here::here("data/analysis_data/cleaned_sex_age_race.parquet"))
+  
+  output$plot <- renderPlot({
+    # Filter data based on user inputs
+    filtered_data <- cleaned_sex_age_race %>%
+      filter(SEX == input$SEX, 
+             AGE %in% input$AGE,
+             RACE %in% input$RACE)
+    
+    ggplot(filtered_data, aes(x = YEAR, y = ESTIMATE, color = RACE, linetype = SEX)) +
+      geom_line() +
+      geom_point() +
+      facet_grid(. ~ AGE) +
+      labs(title = "Suicide Rates between 1950-2018 in United States by Sex, Age and Race",
+           x = "Year", y = "Suicide Rate")
+  })
+}
+
+shinyApp(ui = ui, server = server)

### 2024-04-21 paper/paper.qmd
update

@@
 ---
-title: "Understanding Suicide Rates: Exploring Gender and Age Patterns That Lead to Various Risk Factors Associated with Suicidal Behaviour"
+title: "Exploring Suicide Rates: Assessing Age, Sex and Race Patterns with Statistical Insights into Suicide Risk"
 author: 
@@
 abstract: "This paper utilizes data from Statistics Canada, retrieving data related to the Suicide Rates.
-Analysing the relationship with gender, it is shown that suicide rates tend to be higher in male rather than female.
+Analysing the relationships with sex, age and race it is shown that suicide rates tend to be higher in male rather than female.
 However, suicide rates vary between gender through age, as we find that adolecent females may have a higher rate for suicidal attempts.
@@
 bibliography: references.bib
+editor: 
+  markdown: 
+    wrap: sentence
 ---
@@
 library(tidyverse)
-library(palmerpenguins)
+library(arrow)
+library(knitr)
+library(dplyr)
+library(ggplot2)
+
+```
+
+```{r}
+#| echo: false
+#| eval: true
+#| warning: false
+#| message: false
+#| 
+### Retrieve Model ###
+first_model <- readRDS(here::here("models/first_model.rds"))
+second_model <- readRDS(here::here("models/second_model.rds"))
+third_model <- readRDS(here::here("models/third_model.rds"))
+fourth_model <- readRDS(here::here("models/fourth_model.rds"))
+
 ```
@@
 
-Suicide is a significant worldwide concern as it concerns individuals, friends, families, and communities.
-Understanding the factors of suicide is crucial as it extends beyond a loss of a life and onto ongoing world problems like mental health.
-Hence understanding factors that contribute to suicide is critical into developing prevention strategies within the public health officials and professionals.
+The issue of suicide has gained increasing attention all around the world involving all individuals, friends, families, and communities.
+Suicide rates serve as a critical indicator of the mental health and well-being of a population, reflecting underlying social, economic, and cultural factors.
+Thus, a comprehensive understanding of suicide distribution across demographics are essential for targeted intervention and prevention strategies within the public health officials and professionals.
+However, while existing research acknowledges these disparities, there remains a crucial gap in understanding how specific factors such as age, sex, and race intersect to influence suicide rates.
+Hence the need for specified research that consider each variables and their relationship between how factors play part can further help to come up with answers and measures for a better future.
 
-The complexity of suicide roots from various factors such as vulnerabilities, social determinants, cultural norms, and mental health being the lead cause for suicides.
-Moreover, suicides are varied from different demographic groups such as age and gender.
-Which highlights the need for specified research that consider these variables with suicide.
-Analyzing the relationship between how these factors play part in suicide rates can help further come up with answers and measures for a better future.
+This paper aims to address this gap by delving into the intricate dynamics of suicide rates in United States, focusing specifically on the nuanced intersections of age, sex, and race.
+By employing a linear regression analysis, we seek to unravel how these demographic variables contribute to suicide risk factors.
+The estimand here is how age, gender and race are related to the risk factors of suicides.
+Through this approach, we endeavor to elucidate which demographic groups are at heightened risk of suicide prevalence.
+Drawing upon suicide data sets soured from the "Centers for Disease Control and Prevention (CDC)" [@cdc_suicide_rates], our study unveils temporal patterns and disparities in suicide rates across various demographic groups.
+Through rigorous analysis, we illuminate shifting dynamics in suicide prevalence among different age cohorts, sexes, and racial groups.
+Furthermore, by exploring the intersections of these factors, we reveal how they interact to shape the overall landscape of suicide in the United States.The study focuses on answering the following questions:
 
-In this paper, we aim to explore the issues of suicide, carefully focusing on how age and gender play into its rates through a linear regression.
-The estimand here is how age and gender are related to the risk factors of suicides.
-Meaning, how they play into determining which group of individuals are at high suicidal risk.
-We will draw data from Statistics Canada, specifically focusing in Canadian suicide rates in 2019.
+-   What trend do we see in the total suicide rates in US between 1950-2018?
 
-# Data {#sec-data}
+-   How do suicide rates vary across different age groups?
 
-Some of our data is of penguins (@fig-bills), from @palmerpenguins.
+-   How do suicide rates vary between sexes?
+
+-   How do suicide rates vary across age and sex combined?
+
+-   How do suicide rates vary across sex and race combined?
+
+-   How do suicide rates vary across age, sex, and race?
+
+Understanding the trends and disparities in suicide rates is vital for developing effective prevention and intervention strategies tailored to the diverse needs of vulnerable populations.
+By elucidating the intricate relationships between age, sex, and race in the context of suicide, this paper contributes to the broader discourse on mental health and public health policy.
+In the subsequent sections, we delve deeper into the methodology employed, present our findings, discuss their implications, and propose avenues for future research, thereby offering valuable insights into addressing this pressing public health issue.
+
+## Data Source
+
+The data utilized in this paper concentrates on suicide rates in United States between the years from 1950 to 2018, which is accessible through the database from the Center for Disease Control and Prevention (CDC) on Death rates for Suicide, by sex, race, Hispanic origin, and age: United States [@cdc_suicide_rates].
+This data source contains data concerning suicide rates covering demographics of the following age, race, sex, and Hispanic origin.
+It is available to the public, which can be downloaded for free from its website in forms of multiple export files.
+The data collection process of the database is provided by the National Center for Health Statistics (NCHS) and DAE that samples on a population of 100,000 US residents.
+
+The analysis of this paper makes use of the R programming language [@citeR] for statistical computation and to visualize data.
+The tidyverse package [@rtidyverse] is installed to have access to other R packages used in this paper.
+Other packages used are dplyr [@rdplyr], ggplot2[@rggplot2], knitr [@rknitr], and arrow [@arrow].
+To enhance visualizations, we used the Shiny App [@Rshinyapp] to create a more interactive visualization.
+
+## Variables of Interest
+
+Our analysis is specifically interested in investigating the demographic factors of age, sex, and race between the years 1950 to 2018.
+We have utilized the variables from population of 100,000 US residents crude excluding any age-adjusted populations for a more accurate representation of the population.
+Moreover, cleaning the data in order to separate the sections of the raw data into each selected factors of age, sex, race, and a combination.
+The specific variable in the raw data that was not included in this research are Hispanic origin.
+
+## Total Suicide Rates in United States
+
+The overall trend in Suicide Rates in United States are shown below in [@tbl-totalsuicide].
+Indicating us a percentage of an estimate of deaths per year.
 
 ```{r}
-#| label: fig-bills
-#| fig-cap: Bills of penguins
+#| message: false
+#| echo: false
+#| tbl-cap: Total Suicide Rates per 100,000 population by Year 
+#| label: tbl-totalsuicide
+
+cleaned_total <- read_parquet(here::here("data/analysis_data/cleaned_total.parquet"))
+
+#Create Table
+data <- cleaned_total |>
+  group_by(YEAR) |>
+  summarise(ESTIMATE = sum(ESTIMATE)) |>
+  arrange(desc(YEAR))
+
+knitr::kable(data, 'markdown', col.names = c("Year", "Estimate"), align = 'c')
+
+```
+
+```{r}
+#| label: fig-1
+#| fig-cap: Total Suicide Rates per 100,000 Population by Year
 #| echo: false
 
-ggplot(penguins, aes(x = island, fill = species)) +
-  geom_bar(alpha = 0.8) +
-  scale_fill_manual(values = c("darkorange","purple","cyan4"),
-                    guide = "none") +
-  theme_minimal() +
-  facet_wrap(~species, ncol = 1) +
-  coord_flip()
+cleaned_total <- read_parquet(here::here("data/analysis_data/cleaned_total.parquet"))
+
+cleaned_total |>
+  ggplot(aes(x = YEAR, y = ESTIMATE)) +
+  geom_bar(position = "dodge2", stat = "identity") +
+  labs(x = "Year", y = "Estimate") +
+  theme_minimal()
+
 ```
 
-Talk more about it.
+The plot [@fig-1], shows us a brief understanding of suicide trends by year.
+We can see that from year 2000, there is gradual increase in suicides hitting the highest numbers by 2018.
+[@tbl-totalsuicide] shows us the accurate numbers of suicide rate estimates.
+Providing a estimate of 14.8 estimates in 2018 and the lowest count at 2000 with 10.4 estimation.
+In the following sections we will observe trends of suicide rates by each demographic factors.
 
-And also planes (@fig-planes).
-(You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
+### Trends in Suicide Rates by Sex
+
+There is a significant difference in suicide rates between the two sexes.
+It is evidently provided that males have a higher risk in suicide compared to females shown in [@fig-2].
 
 ```{r}
-#| label: fig-planes
-#| fig-cap: Relationship between wing length and width
+#| label: fig-2
+#| fig-cap: Crude Death Rates per 100,000 Resident Population by Sex
 #| echo: false
-#| warning: false
-#| message: false
 
-analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
+#Retrieve Data
+cleaned_sex <- read_parquet(here::here("data/analysis_data/cleaned_sex.parquet"))
 
-analysis_data |> 
-  ggplot(aes(x = width, y = length)) +
-  geom_point(alpha = 0.8) +
-  theme_minimal() +
-  labs(x = "Wing width (mm)",
-       y = "Wing length (mm)")
+ggplot() + 
+  geom_line(data = cleaned_sex, aes(x = YEAR, y = ESTIMATE, color = factor(SEX), fill = factor((SEX)),stat = "identity")) +
+  geom_line(data = cleaned_total, aes(x = YEAR, y = ESTIMATE, color = POPULATION),
+           stat = "identity") +
+  labs(x = "Year", y = "Death Rate") +
+  scale_fill_brewer(palette = "Set1", name = "Sex") +
+  theme_minimal()
+
 ```
 
-Talk way more about it.
+The estimated baseline suicide rate, represented by the intercept, is approximately 5.0 which indicates the average suicide rates when not considering sex as a factor.
+However, when taking in the effects of sex, the estimated suicide rates for males is approximately 14.3.
+That is 34% more than the average suicide rates.
+This shows us the significant emphasis of the disparities in suicide rates per sex.
+
+### Trends in Suicide Rates by Age
+
+Overall, we see a decrease in suicides in all age ranges excluding 10-14 years from 1950 to 1990s.
+However, from the 1990s, we see a increase into the early 2000s and then again a decreasing until around 2010.
+Then once again an increase from then until 2018.
+The only age group that does not follow this pattern are the 10-14 year range, showing us consistence and a slight increase starting from the early 2000s.
+
+```{r}
+#| label: fig-3
+#| fig-cap: Crude Death Rates per 100,000 Resident Population by Age
+#| echo: false
+
+#Retrieve Data
+cleaned_age <- read_parquet(here::here("data/analysis_data/cleaned_age.parquet"))
+
+#Create graph
+cleaned_age |>
+  ggplot(mapping = aes(x = YEAR, y = ESTIMATE, color = factor(AGE))) +
+  geom_line() +
+  labs(x = "Year", y = "Death Rate", color = "Age Group") +
+  theme_minimal()
+```
+
+Based on the data provided above [@fig-3], we find that apart from ages 10-14, all the other age range show a similiar pattern in suicide rates.
+Analyzing the data furthermore, we see that younger age groups such as those between 15-24 years tend to have lower suicide rates compared to older age groups.
+There is a notable increase in suicide rates among elderly population, particularity those ages 75 year and above.
+When we observe the mean of suicides per age, the lowest mean among the elderly is 18.6 which is higher than younger aged population, highest mean being 12.4.
+Moreover, within each age range, there is a variability in suicide rates that may reflect diverse factors that influence suicide risk within specific age groups.
+Such as socioeconomic status, mental health issues and access to health support.
 
@@
 
-The goal of our modelling strategy is twofold. Firstly,...
-
-Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
+Through the data analysis of this paper, we discovered that there is a correlation between suicide rates and demographic factors such as age, sex, and race.
+To further analyze the implications of the impact demographic factors with suicide rates, we constructed a linear regression model.
 
@@
 
-Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.
+This model fits a linear regression model where year is the independent variable and suicide rates are the response variable.
+We run the model in R [@citeR] using the `rstanarm` package of [@rstanarm].
 
@@
 \begin{align} 
-y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \alpha + \beta_i + \gamma_i\\
-\alpha &\sim \mbox{Normal}(0, 2.5) \\
-\beta &\sim \mbox{Normal}(0, 2.5) \\
-\gamma &\sim \mbox{Normal}(0, 2.5) \\
-\sigma &\sim \mbox{Exponential}(1)
+\y_i = \beta_0 + \beta_1 \mbox{Sex} + \beta_2  \mbox{Age}+ \beta_3 \mbox{Race}
 \end{align}
 ```
-We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.
+Define
 
-### Model justification
+-   $y_i$ as the suicide rate estimated by year
 
-We expect a positive relationship between the size of the wings and time spent aloft. In particular...
+-   $\beta_0$ is the coefficient for intercept
 
-We can use maths by including latex between dollar signs, for instance $\theta$.
+-   $\beta_1$ is the coefficient for the rate at which suicide rates change with respect to sex
+
+-   $\beta_2$ is the coefficient for the rate at which suicide rates change with respect to age
+
+-   $\beta_3$ is the coefficient for the rate at which suicide rates change with respect to race
+
+## Model justification
+
+We anticipate that the linear regression model helps us understand the relationship between suicide rates and several demographic variables such as sex, age, and race.
+This is because each coefficient in the model represents the change in the response variable associated with a one-unit change in the predictor variable (age, sex, and race) that hold other variables constant.
+When using the model, we can determine which demographic factors have a significant impact on suicide rates.
+The significance of each coefficient can be assessed using hypothesis test, such as t-tests.
+Furthermore, when including multiple predictor variables in the model, we can control for potential confounding variable.
+To illustrate, by including both age and sex in the model, we can assess the independent effect of each variable on suicide rates while controlling for the other.
+This model can also capture the interaction effects between predictor variables.
+Demonstrating the effect of age on suicide rates based on the different sexes and racial group.
 
 # Results
+
+## Suicide Rates by Sex and Age
+
+The estimated baseline suicide rates when controlling for sex and age predictors is approximately -9.6.
+This represents the suicide rate when both sexes and age are at their reference levels.
+For males, the estimated effects on suicide rates are approximately 21.5, suggesting that males have significantly higher estimated suicide rates compared to females when other variables are held constant.
+The plot shown below [@fig-4] also suggests the opposite interaction between the years from 1980 to 2018.
+While males are on average showing an concave down pattern, females are presenting a concave up trend.
+
+```{r}
+#| label: fig-4
+#| fig-cap: Suicide Rates per 100,000 Population by Sex and Age
+#| echo: false
+
+cleaned_age_sex <- read_parquet(here::here("data/analysis_data/cleaned_age_sex.parquet"))
+
+# Filter data for males
+male_data <- cleaned_age_sex %>%
+  filter(SEX == "Male", AGE %in% AGE)
+
+# Filter data for females
+female_data <- cleaned_age_sex %>%
+  filter(SEX == "Female", AGE %in% AGE)
+
+# Plot for males
+ggplot(male_data, aes(x = YEAR, y = ESTIMATE, color = AGE, fill = AGE)) +
+  geom_line() +
+  geom_point() +
+  labs(title = "Suicide Rates for Males by Age Groups",
+       x = "Year",
+       y = "Suicide Rate per 100,000 resident population",
+       color = "Age Group",
+       fill = "Age Group") +
+  scale_fill_viridis_d(option = "A", name = "Age") +
+  scale_color_viridis_d(option = "A", name = "Age") +
+  theme(legend.position = "right") +
+  facet_wrap(~SEX)
+
+# Plot for females
+ggplot(female_data, aes(x = YEAR, y = ESTIMATE, color = AGE, fill = AGE)) +
+  geom_line() +
+  geom_point() +
+  labs(title = "Suicide Rates for Females by Age Groups",
+       x = "Year",
+       y = "Suicide Rate per 100,000 resident population",
+       color = "Age Group",
+       fill = "Age Group") +
+  scale_fill_viridis_d(option = "A", name = "Age") +
+  scale_color_viridis_d(option = "A", name = "Age") +
+  theme(legend.position = "right") +
+  facet_wrap(~SEX)
+```
+
+Visibility we see that suicide rates vary across different age groups.
+Generally, suicide rates increase with age, as shown in the previous observation in [@fig-3].
+With older age groups having a higher estimated suicide compared to younger age groups.
+These findings underscore the significance of considering both gender and age when examining suicide rates, as they provide valuable insights into the complex interplay of demographic factors influencing suicide risk.
+Moreover, they highlight the importance of tailored interventions targeting specific demographic groups to prevent suicide effectively.
+
+### Suicide Rates by Sex and Race
+
+Shown below in [@fig-5], we can observe the relationship between sex and race through a bar graph.
+Comparing female and male among the years selected from 1980, 1990, 2000, 2005, 2010, 2015, 2016, 2017, 2018 by racial groups.
+We selected more recent years, to see more relevant data to today.
+
+```{r}
+#| label: fig-5
+#| fig-cap: Crude Death Rates per 100,000 Resident Population by Sex and Race per Selected Year
+#| echo: false
+
+cleaned_sex_race <- read_parquet(here::here("data/analysis_data/cleaned_sex_race.parquet"))
+
+# Choose specific years
+years <- c(1980, 1990, 2000, 2005, 2010, 2015, 2016, 2017, 2018)
+
+# Filter data for selected years
+cleaned_sex_race_selected <- cleaned_sex_race |>
+  filter(YEAR %in% years)
+
+# Create separate bar charts for the selected years
+ggplot(cleaned_sex_race_selected, aes(x = RACE, y = ESTIMATE, fill = SEX)) +
+  geom_bar(position = "dodge2", stat = "identity") +
+  facet_wrap(~ YEAR, scales = "free") +
+  labs(
+    x = "Race",
+    y = "Death Rate",
+    fill = "Sex"
+  ) +
+  theme_minimal() +
+  theme(
+    legend.position = "bottom",
+    axis.text.x = element_text(angle = 40, vjust = 1, hjust = 1),  # Rotate x-axis text
+  ) +
+  scale_fill_brewer(palette = "Set1")
+
+```
+
+The estimated mean response when all predictor variables are zero, in this case, its estimated to be 6.1.
+Focusing on the males, the estimated difference in the response variable between male and female are found that males have a higher response by 10.7.
+We see that individuals part of the Asian or Pacific Islanders and Black or African American races are estimated to have lower responses by 5.4 and 5.3.
+Hence providing us that individuals of the White race are estimated to have a higer response by 2.1 compared to the other racial groups.
+
+### Suicide Rates by Sex, Age, and Race
+
+```{r}
+### ShinyApp for Suicide Rates by Sex, Age and Race ###
+
+# Load required libraries
+library(shiny)
+library(ggplot2)
+
+# Define UI
+ui <- fluidPage(
+  titlePanel("Suicide Rates between 1950-2018 in United States by Sex, Age and Race"),
+  sidebarLayout(
+    sidebarPanel(
+      selectInput("SEX", "Select Sex:", choices = c("Male", "Female")),
+      checkboxGroupInput("AGE", "Select Age Group:", 
+                         choices = c("15-24 years", "25-44 years", "45-64 years", "65 years and over", "75-84 years", "85 years and over"),
+                         selected = c("15-24 years", "25-44 years","45-64 years", "65 years and over", "75-84 years", "85 years and over")),
+      checkboxGroupInput("RACE", "Select Age Group:", 
+                         choices = c("White", "Black or African American", "American Indian or Alaska Native", "Asian or Pacific Islander"),
+                         selected = c("White", "Black or African American", "American Indian or Alaska Native", "Asian or Pacific Islander"))
+    ),
+    mainPanel(
+      plotOutput("plot")
+    )
+  )
+)
+
+server <- function(input, output) {
+
+  # Retrieve data from file
+  cleaned_sex_age_race <- read_parquet(here::here("data/analysis_data/cleaned_sex_age_race.parquet"))
+
+  output$plot <- renderPlot({
+    # Filter data based on user inputs
+    filtered_data <- cleaned_sex_age_race %>%
+      filter(SEX == input$SEX, 
+             AGE %in% input$AGE,
+             RACE %in% input$RACE)
+
+    ggplot(filtered_data, aes(x = YEAR, y = ESTIMATE, color = RACE, linetype = SEX)) +
+      geom_line() +
+      geom_point() +
+      facet_grid(. ~ AGE) +
+      labs(title = "Suicide Rates between 1950-2018 in United States by Sex, Age and Race",
+           x = "Year", y = "Suicide Rate")
+  })
+}
+
+shinyApp(ui = ui, server = server)
+
+```
+
+```{r}
+# Load required libraries
+library(ggplot2)
+library(dplyr)
+
+cleaned_sex_age_race <- read_parquet(here::here("data/analysis_data/cleaned_sex_age_race.parquet"))
+fifth_model <-
+  readRDS(file = here::here("models/fifth_model.rds"))
+
+cleaned_sex_age_race$predicted <- predict(model)
+
+ggplot(cleaned_sex_age_race, aes(x = predicted, y = ESTIMATE)) +
+  geom_point() +
+  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") + # Add a diagonal line for reference
+  labs(title = "Regression Model: Actual vs. Predicted Suicide Rates",
+       x = "Predicted Suicide Rate",
+       y = "Actual Suicide Rate")
+# Print model summary
+summary(fifth_model)
+```
 
@@
 
+1.  **Model Performance**:
+    -   Fit diagnostics, including the mean of the posterior predictive distribution (mean_PPD), suggest that the model adequately predicts suicide rates across genders.
+    -   MCMC diagnostics indicate convergence of the Markov Chain Monte Carlo (MCMC) chains and sufficient effective sample sizes, enhancing confidence in the reliability of the estimates.
+2.  **Policy Implications**:
+    -   The higher estimated suicide rate among males highlights the importance of gender-specific interventions and support services for suicide prevention.
+    -   Tailored strategies addressing the unique risk factors and mental health needs of males can help reduce the burden of suicide in this demographic.
+
+```{=html}
+<!-- -->
+```
+1.  **Policy Implications**:
+    -   These findings underscore the importance of age-specific interventions and support services for suicide prevention.
+    -   Targeted strategies tailored to the unique needs and risk factors of different age groups can help mitigate suicide risk and promote mental well-being across the lifespan.
+
+```{=html}
+<!-- -->
+```
+1.  **Model Performance**:
+
+    -   Fit diagnostics, including the mean of the posterior predictive distribution (mean_PPD), suggest that the model adequately predicts suicide rates considering gender and age as predictors.
+
+    -   MCMC diagnostics indicate convergence of the Markov Chain Monte Carlo (MCMC) chains and sufficient effective sample sizes, enhancing confidence in the reliability of the estimates.**Sigma**: Sigma represents the standard deviation of the residual errors in the model.
+        It indicates the variability of the response variable around the predicted values.
+        In this case, it's estimated to be 2.6.
+
+    The fit diagnostics provide additional information about the model performance:
+
+    -   **mean_PPD**: This represents the mean of the posterior predictive distribution, which is a measure of how well the model predicts the observed data.
+        In this case, it's estimated to be 9.4.
+
+    In summary, these findings indicate the estimated effects of sex and race on the response variable, along with the variability in the model's predictions and residuals.
+
 ## First discussion point {#sec-first-point}
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+If my paper were 10 pages, then should be be at least 2.5 pages.
+The discussion is a chance to show off what you know and what you learnt from all this.
 
@@
 
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...
+In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check.
+This shows...
 
-In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...
+In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior.
+This shows...
 
@@
 
-@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...
+@fig-stanareyouokay-1 is a trace plot.
+It shows...
+This suggests...
 
-@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...
+@fig-stanareyouokay-2 is a Rhat plot.
+It shows...
+This suggests...
 

### 2024-04-21 scripts/02-data_cleaning.R
update

@@ -14,12 +14,22 @@ library(arrow)
 
 
 #### Clean data ####
-raw_data <- read_csv("data/raw_data/Death_rates_for_suicide__by_sex__race__Hispanic_origin__and_age__United_States_20240410.csv")
+raw_data <- read_csv("data/raw_data/Death_rates_for_suicide__by_sex__race__Hispanic_origin__and_age__United_States_20240408.csv")
+
+#UNIT must be by population calculated by crude to determine the un-adjusted rates
+cleaned_total <- raw_data %>%
+  select(UNIT, YEAR, STUB_LABEL, ESTIMATE) %>%
+  filter(STUB_LABEL %in% c("All persons")) %>%
+  filter(UNIT %in% c("Deaths per 100,000 resident population, crude")) %>%
+  rename(POPULATION = STUB_LABEL)
+
+write_parquet(cleaned_total, "data/analysis_data/cleaned_total.parquet")
 
 ###Total suicide rate by sex###
 cleaned_sex <- raw_data %>%
   select(UNIT, YEAR, STUB_LABEL, ESTIMATE) %>%
   filter(STUB_LABEL %in% c("Male", "Female")) %>%
+  filter(UNIT %in% c("Deaths per 100,000 resident population, crude")) %>%
   rename(SEX = STUB_LABEL)
 
 write_parquet(cleaned_sex, "data/analysis_data/cleaned_sex.parquet")
@@ -34,13 +44,28 @@ cleaned_age <- raw_data %>%
 
 write_parquet(cleaned_age, "data/analysis_data/cleaned_age.parquet")
 
+###Total suicide rate by age range and sex###
+cleaned_age_sex <- raw_data %>%
+  select(UNIT, YEAR, STUB_LABEL, ESTIMATE) %>%
+  filter(STUB_LABEL %in% c("Male: 10-14 years", "Male: 15-24 years", "Male: 20-24 years", "Male: 25-44 years", "Male: 35-44 years",
+                           "Male: 45-64 years", "Male: 55-64 years", "Male: 65 years and over", "Male: 65-74 years", "Male: 75-84 years", 
+                           "Male: 85 years and over", "Female: 10-14 years", "Female: 15-24 years", "Female: 20-24 years", "Female: 25-44 years", "Female: 35-44 years",
+                           "Female: 45-64 years", "Female: 55-64 years", "Female:65 years and over", "Female: 65-74 years", "Female: 75-84 years", 
[…]

### 2024-04-21 scripts/04-model.R
update

@@ -1,31 +1,91 @@
 #### Preamble ####
-# Purpose: Models... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Models for Suicide rates based on age, sex and race
+# Author: Sarah Lee
+# Date: 11 April 2024
+# Contact: sarahhhh.lee@mail.utoronto.ca 
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: Parquet files for cleaned suicide data
 
 
 #### Workspace setup ####
 library(tidyverse)
 library(rstanarm)
+library(arrow)
 
 #### Read data ####
-analysis_data <- read_csv("data/analysis_data/analysis_data.csv")
+cleaned_age <- read_parquet("data/analysis_data/cleaned_age.parquet")
+cleaned_sex <- read_parquet("data/analysis_data/cleaned_sex.parquet")
+cleaned_sex_race <- read_parquet("data/analysis_data/cleaned_sex_race.parquet")
+cleaned_sex_age_race <- read_parquet("data/analysis_data/cleaned_sex_age_race.parquet")
 
 ### Model data ####
+
+#Estimate is the suicide rate (dependent variable), creating a model based on age
 first_model <-
   stan_glm(
-    formula = flying_time ~ length + width,
-    data = analysis_data,
+    formula = ESTIMATE ~ AGE,
+    data = cleaned_age,
     family = gaussian(),
     prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
     prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
     prior_aux = exponential(rate = 1, autoscale = TRUE),
     seed = 853
   )
+summary(first_model)
+
+#Estimate is the suicide rate (dependent variable), creating a model based on sex
+second_model <-
+  stan_glm(
+    formula = ESTIMATE ~ SEX,
+    data = cleaned_sex,
+    family = gaussian(),
+    prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
+    prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
+    prior_aux = exponential(rate = 1, autoscale = TRUE),
[…]

### 2024-04-21 scripts/00-simulate_data.R
update

@@ -11,6 +11,10 @@
 #### Workspace setup ####
 library(tidyverse)
 library(dplyr)
+library(dplyr)
+library(waffle)
+
+
 
 #### Simulate data ####
 # Set seed for reproducibility

### 2024-04-21 scripts/03-test_data.R
update

@@ -1,15 +1,74 @@
 #### Preamble ####
-# Purpose: Tests... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Tests for cleaned data
+# Author: Sarah Lee
+# Date: 18 April 2024
+# Contact: sarahhhh.lee@mail.utoronto.ca 
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: cleaned data from the raw data
 
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
+library(arrow)
+library(testthat)
+
+cleaned_total <- read_parquet(here::here("data/analysis_data/cleaned_total.parquet"))
+cleaned_sex <- read_parquet(here::here("data/analysis_data/cleaned_sex.parquet"))
+cleaned_age <- read_parquet(here::here("data/analysis_data/cleaned_age.parquet"))
+cleaned_sex_race <- read_parquet(here::here("data/analysis_data/cleaned_sex_race.parquet"))
+cleaned_age_sex <- read_parquet(here::here("data/analysis_data/cleaned_age_sex.parquet"))
+cleaned_sex_age_race <- read_parquet(here::here("data/analysis_data/cleaned_sex_age_race.parquet"))
+
 
 #### Test data ####
+#cleaned_total
+is.character(cleaned_sex$UNIT)
+is.numeric(cleaned_sex$YEAR)
+is.numeric(cleaned_sex$ESTIMATE)
+
+#cleaned_sex
+is.character(cleaned_sex$UNIT)
+is.numeric(cleaned_sex$YEAR)
+is.character(cleaned_sex$SEX)
+is.numeric(cleaned_sex$ESTIMATE)
+
+#cleaned_age
+# Check if variables are numeric or character
+variable <- names(cleaned_age)
+variable_types <- sapply(cleaned_age, class)
+
+# Print the variable names and their types
+for(i in 1:length(cleaned_age)) {
+  cat("variable:", variable[i], "\t Type:", variable_types[i], "\n")
+}
+
+#cleaned_sex_race
+# Check if variables are numeric or character
+variable <- names(cleaned_sex_race)
+variable_types <- sapply(cleaned_sex_race, class)
+
[…]

### 2024-04-21 paper/paper.qmd
update paper

--- 2024-04-02 paper/paper.qmd
@@
 ---
-title: "Understanding Suicide Rates: Exploring Gender and Age Patterns That Lead to Various Risk Factors Associated with Suicidal Behaviour"
+title: "Exploring Suicide Rates: Assessing Age, Sex and Race Patterns with Statistical Insights into Suicide Risk"
 author: 
   - Sarah Lee
-thanks: "Code and data are available at: https://github.com/sarahhhh02/final-paper.git"
+thanks: "Code and data are available at: https://github.com/sarahhhh02/suicides.git"
 date: today
 date-format: long
-abstract: "This paper utilizes data from Statistics Canada, retrieving data related to the Suicide Rates.
-Analysing the relationship with gender, it is shown that suicide rates tend to be higher in male rather than female.
-However, suicide rates vary between gender through age, as we find that adolecent females may have a higher rate for suicidal attempts.
-The results of these findings aids policymakers, public health officials, and mental health professionals to develop a more effective and strategic suicide prevention tailored for the population."
+abstract: "This paper investigates the intricate dynamics of suicide rates in the United States, with specific focus on how age, sex, and race intersects.
+Utilizing linear regression analysis and data from the Centers for Disease Control and Prevention (CDC), we unveil temporal patterns and disparities in suicide rates acrross demographic groups.
+Our analysis reveals that suicide rates tend to be higher among elderly male individuals within the white racial group; however, we also observe variations across intersections of age, sex and race.
+These findings offer valuable insights for policymakers, public health officials, and mental health professionals, facillitating the development of more targeted and strategic suicide prevention tailored to the diverse needs of the suicide population."
 format: pdf
@@
 library(tidyverse)
-library(palmerpenguins)
+library(arrow)
+library(knitr)
+library(dplyr)
+library(ggplot2)
+library(rstanarm)
+
 ```
-
-# Introduction
-
-Suicide is a significant worldwide concern as it concerns individuals, friends, families, and communities.
-Understanding the factors of suicide is crucial as it extends beyond a loss of a life and onto ongoing world problems like mental health.
-Hence understanding factors that contribute to suicide is critical into developing prevention strategies within the public health officials and professionals.
-
-The complexity of suicide roots from various factors such as vulnerabilities, social determinants, cultural norms, and mental health being the lead cause for suicides.
-Moreover, suicides are varied from different demographic groups such as age and gender.
-Which highlights the need for specified research that consider these variables with suicide.
-Analyzing the relationship between how these factors play part in suicide rates can help further come up with answers and measures for a better future.
-
-In this paper, we aim to explore the issues of suicide, carefully focusing on how age and gender play into its rates through a linear regression.
-The estimand here is how age and gender are related to the risk factors of suicides.
-Meaning, how they play into determining which group of individuals are at high suicidal risk.
-We will draw data from Statistics Canada, specifically focusing in Canadian suicide rates in 2019.
-
-# Data {#sec-data}
-
-Some of our data is of penguins (@fig-bills), from @palmerpenguins.
-
-```{r}
-#| label: fig-bills
-#| fig-cap: Bills of penguins
-#| echo: false
-
-ggplot(penguins, aes(x = island, fill = species)) +
-  geom_bar(alpha = 0.8) +
-  scale_fill_manual(values = c("darkorange","purple","cyan4"),
-                    guide = "none") +
-  theme_minimal() +
-  facet_wrap(~species, ncol = 1) +
-  coord_flip()
-```
-
-Talk more about it.
-
-And also planes (@fig-planes).
-(You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
-
-```{r}
-#| label: fig-planes
-#| fig-cap: Relationship between wing length and width
-#| echo: false
-#| warning: false
-#| message: false
-
-analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
-
-analysis_data |> 
-  ggplot(aes(x = width, y = length)) +
-  geom_point(alpha = 0.8) +
-  theme_minimal() +
-  labs(x = "Wing width (mm)",
-       y = "Wing length (mm)")
-```
-
-Talk way more about it.
-
-# Model
-
-The goal of our modelling strategy is twofold. Firstly,...
-
-Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
-
-## Model set-up
-
-Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.
-
-\begin{align} 
-y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \alpha + \beta_i + \gamma_i\\
-\alpha &\sim \mbox{Normal}(0, 2.5) \\
-\beta &\sim \mbox{Normal}(0, 2.5) \\
-\gamma &\sim \mbox{Normal}(0, 2.5) \\
-\sigma &\sim \mbox{Exponential}(1)
-\end{align}
-
-We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.
-
-### Model justification
-
-We expect a positive relationship between the size of the wings and time spent aloft. In particular...
-
-We can use maths by including latex between dollar signs, for instance $\theta$.
-
-# Results
-
-Our results are summarized in @tbl-modelresults.
 
@@
 #| message: false
+#| 
+### Retrieve Model ###
+#For statistical analysis on age, sex, age_sex, race_sex
+first_model <- readRDS(here::here("models/first_model.rds"))
+second_model <- readRDS(here::here("models/second_model.rds"))
+third_model <- readRDS(here::here("models/third_model.rds"))
+fourth_model <- readRDS(here::here("models/fourth_model.rds"))
 
-library(rstanarm)
+```
 
-first_model <-
-  readRDS(file = here::here("models/first_model.rds"))
+# Introduction
+
+The issue of suicide has garnered increasied global attention, affecting individuals, friends, families, and communities worldwide.
+Suicide rates serve as a critical indicator of a population's mental health and well-being, reflecting underlying social, economic, and cultural factors.
+Therefore, gaining a comprehensive understanding of suicide distribution across demographics is essential for targeted intervention and prevention strategies by the public health officials and professionals.
+However, while existing research acknowledges these disparities, a critical gap remains in understanding how specific factors intersect to influence suicide rates.
+Hence there is a need for specified research that consider each variables and their relationship to further elucidate answers and measures for a better future.
+
+This paper aims to address this gap by examining the complex dynamics of suicide rates in the United States, focusing specifically on the nuanced intersections of age, sex, and race.
+Through the use of linear regression analysis, we aim to uncover how these demographic variables contribute to suicide risk factors.The estimand here is how age, gender and race are related to the risk factors of suicides.
+Our goal is to understand how age, sex, and race are related to suicide risk factors.
+Through this approach, we seek to identify which demographic groups are at heightened risk of suicide prevalence.
+
+Drawing upon suicide datasets soured from the Centers for Disease Control and Prevention (CDC) [@cdc_suicide_rates], our study reveals temporal patterns and disparities in suicide rates across various demographic groups.
+By rigorous analyzing the data, we illuminate shifting dynamics in suicide prevalence among different age cohorts, sexes, and racial groups.
+Furthermore, by exploring the intersections of these factors, we reveal how they interact to shape the overall landscape of suicide in the United States.
+
+The study focuses on answering the following questions:
+
+-   What trend do we see in the total suicide rates in US between 1950-2018?
+
+-   How do suicide rates vary across different age groups?
+
+-   How do suicide rates vary between sexes?
+
+-   How do suicide rates vary across age and sex combined?
+
+-   How do suicide rates vary across sex and race combined?
+
+-   How do suicide rates vary across age, sex, and race?
+
+Understanding the trends and disparities in suicide rates is crucial for developing effective prevention and intervention strategies tailored to the diverse needs of vulnerable populations.
+By elucidating the intricate relationships between age, sex, and race in the context of suicide, this paper contributes to the broader discourse on mental health and public health policy.
+In the subsequent sections, we delve deeper into the methodology employed, present our findings, discuss their implications, and propose avenues for future research, thereby offering valuable insights into addressing this pressing public health issue.
+
+# Data
+
+## Data Source
+
+The data utilized in this paper focuses on suicide rates in United States from from 1950 to 2018, sourced from the CDC's database on Death rates for Suicide, categorized by sex, race, Hispanic origin, and age [@cdc_suicide_rates].
+This publically acessible data source contains information concerning suicide rates across various demographics, and can be freely downloaded from the CDC's website in multiple export file formats.
+The database is compiled by the National Center for Health Statistics (NCHS) and DAE, sampling from a population of 100,000 US residents.
+
+For the analysis of this paper, we utilized the R programming language [@citeR] for statistical computation and data visualization.
+The tidyverse package[@rtidyverse] was installed to access to other R packages necessary for this paper's analysis.
+Additional packages used include dplyr [@rdplyr], ggplot2 [@rggplot2], knitr [@rknitr], rstanarm [@rstanarm] and arrow [@arrow].
+
+## Variables of Interest
+
+Our analysis focuses specifically on investigating the demographic factors of age, sex, and race between the years 1950 to 2018.
+We utilized data from a population of 100,000 US residents, excluding any age-adjusted populations to ensure a more accurate representation of the population.
+Furthermore, we cleaned the data to separate the sections of the raw data into each selected factors of age, sex, race, and their combination.
+It is worth noting that Hispanic origin was not included as a specific variable in this research.
+
+# Model
+
+Through the data analysis in this paper, we identified correlations between suicide rates and demographic factors such as age, sex, and race.
+To further analyze the implications of these demographic factors with suicide rates, we constructed a linear regression model.
+
+## Model set-up
+
+This model fits a linear regression model where year is the independent variable and suicide rates are the response variable. We run the model in R @citeR using the `rstanarm` package of [@rstanarm].
+
+```{=tex}
+\begin{align} 
+y_i = \beta_0 + \beta_1 \mbox{Sex} + \beta_2  \mbox{Age}+ \beta_3 \mbox{Race}
+\end{align}
 ```
+Define
+
+-   $y_i$ as the suicide rate estimated by year
+
+-   $\beta_0$ is the coefficient for intercept
+
+-   $\beta_1$ is the coefficient for the rate at which suicide rates change with respect to sex
+
+-   $\beta_2$ is the coefficient for the rate at which suicide rates change with respect to age
+
+-   $\beta_3$ is the coefficient for the rate at which suicide rates change with respect to race
+
+## Model justification
+
+We anticipate that the linear regression model will aid in understanding the relationship between suicide rates and various demographic variables such as sex, age, and race.
+Each coefficient in the model represents the change in the response variable associated with a one-unit change in the predictor variable (age, sex, and race), while holding other variables constant.
+By utilizing this model, we can determine which demographic factors have a significant impact on suicide rates.
+The significance of each coefficient can be assessed using hypothesis test, such as t-tests.
+
+Moreover, when incorporating multiple predictor variables in the model, we can control for potential confounding variables.
+To illustrate, by including both age and sex in the model, we can assess the independent effect of each variable on suicide rates while controlling for the other.
+This model can also capture the interaction effects between predictor variables., demonstrating the influence of age on suicide rates based on the different sexes and racial group.
+
+# Results
+
+## Total Suicide Rates in United States
+
+The overall trend in Suicide Rates in United States are shown below in @tbl-totalsuicide. Indicating us an estimated number of suicide deaths per year.
+
+```{r}
+#| message: false
+#| echo: false
+#| tbl-cap: Total Suicide Rates per 100,000 population by Year 
+#| label: tbl-totalsuicide
+
+cleaned_total <- read_parquet(here::here("data/analysis_data/cleaned_total.parquet"))
+
+#Create Table
+data <- cleaned_total |>
+  group_by(YEAR) |>
+  summarise(ESTIMATE = sum(ESTIMATE)) |>
+  arrange(desc(YEAR))
+
+knitr::kable(data, 'markdown', col.names = c("Year", "Estimate"), align = 'c')
+
+```
+
+```{r}
+#| label: fig-1
+#| fig-cap: Total Suicide Rates per 100,000 Population by Year
+#| echo: false
+
+cleaned_total <- read_parquet(here::here("data/analysis_data/cleaned_total.parquet"))
+
+cleaned_total |>
+  ggplot(aes(x = YEAR, y = ESTIMATE)) +
+  geom_bar(position = "dodge2", stat = "identity") +
+  labs(x = "Year", y = "Estimate") +
+  theme_minimal()
+
+```
+
+The plot in @fig-1, provides a snapshot of suicide trends by year.
+We observe a gradual increase in suicides from the year 2000, with the highest numbers recorded by 2018.
+@tbl-totalsuicide presents the accurate numbers of suicide rate estimates, indicating an estimate of 14.8 suicides per 100,000 individuals in 2018, and the lowest count recorded in 2000 with 10.4 estimations.
+In the subsequent sections, we will delve into the trends of suicide rates according to each demographic factors.
+
+## Trends in Suicide Rates by Sex
+
+There is a notable disparity in suicide rates between the two sexes as illustrated in @fig-2. Males exhibit a significant higher risk of suicide compared to females.
+
+```{r}
+#| label: fig-2
+#| fig-cap: Crude Death Rates per 100,000 Resident Population by Sex
+#| echo: false
+
+#Retrieve Data
+cleaned_sex <- read_parquet(here::here("data/analysis_data/cleaned_sex.parquet"))
+
+ggplot() + 
+  geom_line(data = cleaned_sex, aes(x = YEAR, y = ESTIMATE, color = SEX, fill = SEX), stat = "identity") +
+  geom_line(data = cleaned_total, aes(x = YEAR, y = ESTIMATE, color = POPULATION),
+           stat = "identity") +
+  labs(x = "Year", y = "Death Rate") +
+  scale_fill_brewer(palette = "Set1", name = "Sex") +
+  theme_minimal()
+
+```
+
+The estimated baseline suicide rate, represented by the intercept, is approximately 5.0, indicating the average suicide rates when not considering sex as a factor.
+However, when factoring in the effects of sex, the estimated suicide rates for males rises to approximately 14.3.
+That represents 34% increase compared to the average suicide rate.
+This highlights the significant emphasis on the disparities in suicide rates between sexes.
+
+## Trends in Suicide Rates by Age
+
+Overall, we observe a decline in suicides across all age ranges from 1950 to 1990s, except for the 10-14 age group.
+However, starting from the 1990s, we notice an increase in suicides extending into the early 2000s, followed by a decrease until around 2010.
+Subsequently, there is another increase from then until 2018.
+The 10-14 age group stands out from this pattern, exhibiting consistency with a slight increase starting from the early 2000s.
+
+```{r}
+#| label: fig-3
+#| fig-cap: Crude Death Rates per 100,000 Resident Population by Age
+#| echo: false
+
+#Retrieve Data
+cleaned_age <- read_parquet(here::here("data/analysis_data/cleaned_age.parquet"))
+
+#Create graph
+cleaned_age |>
+  ggplot(mapping = aes(x = YEAR, y = ESTIMATE, color = factor(AGE))) +
+  geom_line() +
+  labs(x = "Year", y = "Death Rate", color = "Age Group") +
+  theme_minimal()
+```
+
+Based on the data provided in @fig-3, we observe that, apart from ages 10-14, all other age ranges show a similar pattern in suicide rates.
+Further analysis reveals that younger age groups, such as those between 15-24 years, tend to have lower suicide rates compared to older age groups.
+However, there is a notable increase in suicide rates among the elderly population, particularity those aged 75 year and above.
+
+When examining the mean of suicides per age group, we find that the lowest mean among the elderly is 18.6 which is higher than of younger age groups, with the highest mean being 12.4.
+Moreover, within each age range, there is a variability in suicide rates that may reflect diverse factors that influence suicide risk within specific age groups.
+these factors include socioeconomic status, mental health issues and access to health support.
+
+## Relation to Suicide Rates by Sex and Age
+
+The estimated baseline suicide rates, when controlling for sex and age predictors, is approximately -9.6.
+This represents the suicide rate when both sexes and age are at their reference levels.
+For males, the estimated effects on suicide rates are approximately 21.5, suggesting that males have significantly higher estimated suicide rates compared to females when other variables are held constant.
+
+Furthermore, the plot shown in @fig-4 suggests an opposite interaction between the years from 1980 to 2018. While males on average exhibit a concave down pattern, females present a concave up trend.
+
+```{r}
+#| label: fig-4
+#| fig-cap: Suicide Rates per 100,000 Population by Sex and Age
+#| echo: false
+
+cleaned_age_sex <- read_parquet(here::here("data/analysis_data/cleaned_age_sex.parquet"))
+
+# Filter data for males
+male_data <- cleaned_age_sex %>%
+  filter(SEX == "Male", AGE %in% AGE)
+
+# Filter data for females
+female_data <- cleaned_age_sex %>%
+  filter(SEX == "Female", AGE %in% AGE)
+
+# Plot for males
+ggplot(male_data, aes(x = YEAR, y = ESTIMATE, color = AGE, fill = AGE)) +
+  geom_line() +
+  geom_point() +
+  labs(x = "Year",
+       y = "Suicide Rate per 100,000 resident population",
+       color = "Age Group",
+       fill = "Age Group") +
+  scale_fill_viridis_d(option = "A", name = "Age") +
+  scale_color_viridis_d(option = "A", name = "Age") +
+  theme(legend.position = "right") +
+  facet_wrap(~SEX)
+
+# Plot for females
+ggplot(female_data, aes(x = YEAR, y = ESTIMATE, color = AGE, fill = AGE)) +
+  geom_line() +
+  geom_point() +
+  labs(x = "Year",
+       y = "Suicide Rate per 100,000 resident population",
+       color = "Age Group",
+       fill = "Age Group") +
+  scale_fill_viridis_d(option = "A", name = "Age") +
+  scale_color_viridis_d(option = "A", name = "Age") +
+  theme(legend.position = "right") +
+  facet_wrap(~SEX)
+```
+
+Visibility we see that suicide rates vary across different age groups.
+Generally, suicide rates increase with age, as previously observed in @fig-3, with older age groups exhibiting a higher estimated suicides compared to younger age groups.These findings underscore the importance of considering both gender and age when examining suicide rates, as they offer valuable insights into the complex interplay of demographic factors influencing suicide risk.
+Moreover, they highlight the importance of tailored interventions targeting specific demographic groups to effectively prevent suicide.
+
+## Relation to Suicide Rates by Sex and Race
+
+Displayed below in @fig-5, we can observe the relationship between sex and race through a bar graph.
+the graph compares females and males among the selected years from 2000, 2010, and 2017 across different racial groups.
+
+```{r}
+#| label: fig-5
+#| fig-cap: Suicide Rates per 100,000 Resident Population by Sex and Race 
+#| echo: false
+
+cleaned_sex_race <- read_parquet(here::here("data/analysis_data/cleaned_sex_race.parquet"))
+
+# Choose specific years
+years <- c(2000, 2010, 2017)
+
+# Filter data for selected years
+cleaned_sex_race_selected <- cleaned_sex_race |>
+  filter(YEAR %in% years)
+
+# Create separate bar charts for the selected years
+ggplot(cleaned_sex_race_selected, aes(x = RACE, y = ESTIMATE, fill = SEX)) +
+  geom_bar(position = "dodge2", stat = "identity") +
+  facet_wrap(~ YEAR, scales = "free") +
+  labs(
+    x = "Race",
+    y = "Death Rate",
+    fill = "Sex"
+  ) +
+  theme_minimal() +
+  theme(
+    legend.position = "bottom",
+    axis.text.x = element_text(angle = 40, vjust = 1, hjust = 1), 
+    strip.text = element_text(margin = margin(b = 0.1))  
+  ) +
+  scale_fill_brewer(palette = "Set1")
+
+```
+
+The estimated mean response when all predictor variables are zero, is 6.1.
+Specifically focusing on males, the estimated difference in the response variable between males and females are found to be 10.7, indicating that males have a higher response by this amount compared to females.
+Additionally, individuals identified as Asian or Pacific Islanders and Black or African American races are estimated to have lower responses by 5.4 and 5.3, respectively.
+This suggests that individuals of the White race are estimated to have a higher response by 2.1 compared to the other racial groups.
+Visually, American Indian or Alaska Native individuals are depicted as having the second highest response, as represented in the graph.
+
+## Relation to Suicide Rates in Sex, Age, and Race
+
+Analyzing suicide rates in the United States over the decades, we observe fluctuations based on the combination of race, age and sex.
+these fluctuations reveal both increases and decreases influenced various factors such as changes in societal attitudes towards mental health, economic conditions, access to healthcare, and public health initiatives.
+Below, in @fig-6, we visualize age, sex, and race in one graph representing suicide rates.
+
+```{r}
+#| label: fig-6
+#| fig-cap: Suicide Rates per 100,000 Resident Population by Sex, Age, and Race
+#| echo: false
+
+cleaned_sex_age_race <- read_parquet(here::here("data/analysis_data/cleaned_sex_age_race.parquet"))
+
+age = c("15-24 years", "25-44 years", "45-64 years", "65 years and over", "75-84 years", "85 years and over")
+
+filtered_data <- cleaned_sex_age_race |>
+  filter(AGE %in% age)
+
+ggplot(filtered_data, aes(x = YEAR, y = ESTIMATE, color = RACE)) +
+  geom_line() +
+  facet_grid(SEX ~ AGE, switch = "y") +
+  labs(title = "Suicide Rates between 1950-2018 in United States by Sex, Age and Race",
+       x = "Year", y = "Suicide Rate") +
+  theme(
+    legend.position = "bottom",
+    axis.text.x = element_text(angle = 40, vjust = 1, hjust = 1),  # Rotate x-axis text
+  )
+
+```
+
+When examining suicide rates in the United States based on sex, age, and race alone, notable trends emerge.
+Female suicide rates exhibit a degree of consistency over time, while male rates demonstrate variability across different age brackets and racial backgrounds.
+It is evident that males consistently show a higher suicide rate compared to females.
+Age serves as a significant factor influencing suicide rates, with historical data indicating a prevalence of suicide among older adults, particularly those ages 65 and above.
+Nonetheless, recent years have witnessed a concerning uptick in suicide rates among younger demographics, notably young adults.
+
+Racial disparities also play a role in shaping suicide rates.
+Historically, White individuals have exhibited higher suicide rates compared to other racial groups in the United States.
+However, there has been a disproportionate rise in suicide rates among certain racial minority groups, such as Native Americans and Alaska Natives, in recent years.
 
@@
 
+fifth_model <- readRDS(file = here::here("models/fifth_model.rds"))
+
 modelsummary::modelsummary(
   list(
-    "First model" = first_model
+    "Model" = fifth_model
   ),
@@
 
+In our investigation of suicide rates in the United States, we utilized a regression model to discern significant patterns across demographic factors such as age, sex, and race.
+The summary in @tbl-modelresults provides intriguing insights into the relationship between age and the dependent variable.
+
+Notably, individuals aged 65 years and over exhibit a substantial increase in the dependent variable, with coefficients of 2.59.
+However, this effect varies across different age brackets within this group.
+While individuals aged 65-74 years display a decrease in the dependent variable, those aged 75-84 years and 85 years and over demonstrate significant increases of 11.23 and 33.91, respectively.
+These findings underscore the nuanced impact of age on the outcome, with older age groups exhibiting divergent associations.
+Another key demographic variable examined in the model is sex.
+The coefficient associated with being male is 14.02, indicating a notable increase in the dependent variable compared to females.
+This finding suggests a potential gender-based disparity in the outcome, warranting further investigation into the underlying factors driving this divergence.
+The model also sheds light on the influence of race on the dependent variable.
+Notably, individuals belonging to Asian or Pacific Islander and Black or African American racial groups exhibit negative coefficients of -9.32 and -10.99, respectively.
+In contrast, individuals identified as White display a positive coefficient of 0.84.
+These findings highlight disparities in the outcome across different racial groups, emphasizing the importance of addressing racial equity in various domains.
+
+Assessing the overall fit of the model, the R-squared value of 0.809 indicates that a substantial proportion of the variability in the dependent variable is explained by the demographic factors included in the model.
+Furthermore, measures such as the adjusted R-squared, log likelihood, and root mean square error provide additional insights into the model's predictive performance and goodness of fit.
+
+In conclusion, the findings of the statistical model underscore the significance of demographic factors in shaping outcomes of interest.
+Age, sex, and race emerge as important predictors, each exhibiting distinct associations with the dependent variable.
+They provide valuable insights into the complex factors that contribute to suicide risk and resilience in the United States.
+Various fields, including healthcare, sociology, and public policy, highlighting the importance of considering demographic diversity in research and decision-making processes.
+
 # Discussion
 
-## First discussion point {#sec-first-point}
+This paper presents a comphrehensive statistical model of the relationship between demographic factors - age, sex, and race - and their impact on suicide rates.
+Our objective is to uncover how these variables contribute to variations in the outcome under study.
+Initially, we conducted separate analysis for each intersection on the demographic factors, we made initial observant for each factor alone.
+Concluding how demographic factors to discern their individual influence on suicide rates .
+This provides understanding into how age and sex intersect to shape suicide risks.
+Subsequently, we explored intersections within the demographic factors to identify any changes in variation.
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+One significant insight gleaned from our analysis is the varying impact of age on the dependent variable.
+Older age groups, notably those aged 75-84 years and 85 years and over, exhibit significant increases in the outcome, emphasizing the critical role of age as a predictor in understanding suicide risk.
+Additionally, our analysis highlights the substantial increase in the dependent variable among males compared to females, indicating potential gender-based disparities that necessitate further exploration and intervention.
+Overall, our findings suggest that White male individuals among other age groups are at risk of suicide rates.
+These observations underscore the multifaceted nature of suicide risk and emphasize the importance of tailored interventions that address the unique needs of diverse demographic groups.
+The intersection of race, age, and sex can significantly influence suicide risk, with older men from certain racial minorities facing distinct challenges and risk factors.
 
-## Second discussion point
-
-## Third discussion point
+Effective suicide prevention strategies should prioritize targeted support systems, mental health resources, and initiatives aimed at reducing stigma and promoting well-being across all segments of society.
+Socioeconomic factors, including poverty, unemployment, social isolation, and access to healthcare, may contribute to suicide risk factors across demographic groups, underscoring the need for comprehensive approaches to address these underlying determinants.
 
@@
 
-Weaknesses and next steps should also be included.
+Despite these insights, several weaknesses in our study are apparent.
+Firstly, while our model provides valuable associations between demographic factors and the outcome, it may lack sufficient granularity to capture the full complexity of these relationships.
+Additionally, the reliance on cross-sectional data limits our ability to establish causality or account for temporal dynamics.
+Moving forward, future research should strive to address these limitations and expand upon our findings to develop targeted interventions aimed at promoting equity and inclusivity across diverse populations.
+
+Longitudinal studies could provide insights into the temporal trajectories of demographic influences on the outcome, allowing for a more nuanced understanding of these relationships over time.
+Moreover, incorporating additional variables, such as socioeconomic status or health behaviors, could enhance the model's explanatory power and predictive accuracy.
+Advanced statistical techniques, such as structural equation modeling or machine learning algorithms, could facilitate the identification of latent patterns and interactions among demographic factors and the outcome.
+
+Qualitative research methods, such as interviews or focus groups, could complement quantitative analyses by providing deeper insights into the lived experiences and perspectives of individuals within different demographic groups.
+In conclusion, while our paper offers valuable insights into the relationship between demographic factors and suicide rates, there remain opportunities for future research to deepen our understanding of these complex phenomena and inform evidence-based interventions aimed at promoting equity and inclusivity in society.
 
 \newpage
-
-\appendix
-
-# Appendix {-}
-
-# Additional data details
-
-# Model details {#sec-model-details}
-
-## Posterior predictive check
-
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...
-
-In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...
-
-```{r}
-#| eval: true
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-ppcheckandposteriorvsprior
-#| layout-ncol: 2
-#| fig-cap: "Examining how the model fits, and is affected by, the data"
-#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
-
-pp_check(first_model) +
-  theme_classic() +
-  theme(legend.position = "bottom")
-
-posterior_vs_prior(first_model) +
-  theme_minimal() +
-  scale_color_brewer(palette = "Set1") +
-  theme(legend.position = "bottom") +
-  coord_flip()
-```
-
-## Diagnostics
-
-@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...
-
-@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...
-
-```{r}
-#| echo: false
-#| eval: true
-#| message: false
-#| warning: false
-#| label: fig-stanareyouokay
-#| fig-cap: "Checking the convergence of the MCMC algorithm"
-#| fig-subcap: ["Trace plot", "Rhat plot"]
-#| layout-ncol: 2
-
-plot(first_model, "trace")
-
-plot(first_model, "rhat")
-```
 

### 2024-04-21 suicide_sex_age_race/app.R
update

@@ -1,55 +1,0 @@
-#
-# This is a Shiny web application. You can run the application by clicking
-# the 'Run App' button above.
-#
-# Find out more about building applications with Shiny here:
-#
-#    https://shiny.posit.co/
-#
-
-# Load required libraries
-library(shiny)
-library(ggplot2)
-
-
-# Define UI
-ui <- fluidPage(
-  titlePanel("Suicide Rates by Sex, Age and Race"),
-  sidebarLayout(
-    sidebarPanel(
-      selectInput("SEX", "Select Sex:", choices = c("Male", "Female")),
-      checkboxGroupInput("AGE", "Select Age Group:", 
-                         choices = c("15-24 years", "25-44 years", "45-64 years", "65 years and over", "75-84 years", "85 years and over"),
-                         selected = c("15-24 years", "25-44 years","45-64 years", "65 years and over", "75-84 years", "85 years and over")),
-      checkboxGroupInput("RACE", "Select Age Group:", 
-                         choices = c("White", "Black or African American", "American Indian or Alaska Native", "Asian or Pacific Islander"),
-                         selected = c("White", "Black or African American", "American Indian or Alaska Native", "Asian or Pacific Islander"))
-    ),
-    mainPanel(
-      plotOutput("plot")
-    )
-  )
-)
-
-server <- function(input, output) {
-  
-  # Retrieve data from file
-  cleaned_sex_age_race <- read_parquet(here::here("data/analysis_data/cleaned_sex_age_race.parquet"))
-  
-  output$plot <- renderPlot({
-    # Filter data based on user inputs
-    filtered_data <- cleaned_sex_age_race %>%
-      filter(SEX == input$SEX, 
-             AGE %in% input$AGE,
-             RACE %in% input$RACE)
-    
-    ggplot(filtered_data, aes(x = YEAR, y = ESTIMATE, color = RACE, linetype = SEX)) +
-      geom_line() +
-      geom_point() +
-      facet_grid(. ~ AGE) +
-      labs(title = "Suicide Rates between 1950-2018 in United States by Sex, Age and Race",
-           x = "Year", y = "Suicide Rate")
-  })
-}
-
-shinyApp(ui = ui, server = server)

### 2024-04-22 paper/paper.qmd
update paper/paper.qmd

@@
 
-This paper presents a comphrehensive statistical model of the relationship between demographic factors - age, sex, and race - and their impact on suicide rates.
+This paper presents a comprehensive statistical model of the relationship between demographic factors - age, sex, and race - and their impact on suicide rates.
 Our objective is to uncover how these variables contribute to variations in the outcome under study.
@@
 Subsequently, we explored intersections within the demographic factors to identify any changes in variation.
+All estimations were done through modelling that can be found in the file path scripts/04-model.R using the model summary of each respective findings.
 

### 2024-04-22 other/datasheet/datasheet_template.qmd
update datasheet

---
title: "Datasheet for 'A dataset'"
author: 
  - Sarah Lee
thanks: "Code and data are available at: https://github.com/sarahhhh02/suicides.git"
date: "`r Sys.time()`"
date-format: "D MMMM YYYY"
format: pdf
number-sections: true
bibliography: references.bib
---

Extract of the questions from @gebru2021datasheets.

**Motivation**

1. *For what purpose was the dataset created? Was there a specific task in mind? Was there a specific gap that needed to be filled? Please provide a description.*
    - The dataset was created to enable analysis of suicide rates in United States. We were unable to find a publicly available dataset in a structured format that had separate columns for demographic factors.
2. *Who created the dataset (for example, which team, research group) and on behalf of which entity (for example, company, institution, organization)?*
    - Centers for Disease Control and Prevention (CDC), National Center for Health Statistics (NCHS), and DAE
3. *Who funded the creation of the dataset? If there is an associated grant, please provide the name of the grantor and the grant name and number.*
    - Centers for Disease Control and Prevention (CDC)
4. *Any other comments?*
    - None

**Composition**

1. *What do the instances that comprise the dataset represent (for example, documents, photos, people, countries)? Are there multiple types of instances (for example, movies, users, and ratings; people and interactions between them; nodes and edges)? Please provide a description.*
	- The instances in the dataset represent death rates for suicide categorized by age, sex, race, and Hispanic origin within the United States population.
2. *How many instances are there in total (of each type, if appropriate)?*
	- Total of 6391 rows
3. *Does the dataset contain all possible instances or is it a sample (not necessarily random) of instances from a larger set? If the dataset is a sample, then what is the larger set? Is the sample representative of the larger set (for example, geographic coverage)? If so, please describe how this representativeness was validated/verified. If it is not representative of the larger set, please describe why not (for example, to cover a more diverse range of instances, because instances were withheld or unavailable).*
	- Covers a 100,000 Resident population in US. However, it's unclear whether this represents a sample or the entire population. 
4. *What data does each instance consist of? "Raw" data (for example, unprocessed text or images) or features? In either case, please provide a description.*
	- Not cleaned. All subsections/coloums are included as rows
5. *Is there a label or target associated with each instance? If so, please provide a description.*
	- The presence of a label or target associated with each instance isn't specified.
6. *Is any information missing from individual instances? If so, please provide a description, explaining why this information is missing (for example, because it was unavailable). This does not include intentionally removed information, but might include, for example, redacted text.*
	- Information regarding missing data or the completeness of individual instances isn't provided. But at glance it seems not all age groups are present when intersecting with race or sex. 
7. *Are relationships between individual instances made explicit (for example, users' movie ratings, social network links)? If so, please describe how these relationships are made explicit.*
	- No mentioned
8. *Are there recommended data splits (for example, training, development/validation, testing)? If so, please provide a description of these splits, explaining the rationale behind them.*
	- None
9. *Are there any errors, sources of noise, or redundancies in the dataset? If so, please provide a description.*
	- Not sure
10. *Is the dataset self-contained, or does it link to or otherwise rely on external resources (for example, websites, tweets, other datasets)? If it links to or relies on external resources, a) are there guarantees that they will exist, and remain constant, over time; b) are there official archival versions of the complete dataset (that is, including the external resources as they existed at the time the dataset was created); c) are there any restrictions (for example, licenses, fees) associated with any of the external resources that might apply to a dataset consumer? Please provide descriptions of all external resources and any restrictions associated with them, as well as links or other access points, as appropriate.*
	- Relies on the original data collector (NCHS)
11. *Does the dataset contain data that might be considered confidential (for example, data that is protected by legal privilege or by doctor-patient confidentiality, data that includes the content of individuals' non-public communications)? If so, please provide a description.*
	- None
12. *Does the dataset contain data that, if viewed directly, might be offensive, insulting, threatening, or might otherwise cause anxiety? If so, please describe why.*
	- Possibly as it include sex and race
13. *Does the dataset identify any sub-populations (for example, by age, gender)? If so, please describe how these subpopulations are identified and provide a description of their respective distributions within the dataset.*
	- The dataset likely identifies sub-populations based on age, gender, race, and Hispanic origin, but specific details about how these subpopulations are identified and distributed within the dataset are not provided.
14. *Is it possible to identify individuals (that is, one or more natural persons), either directly or indirectly (that is, in combination with other data) from the dataset? If so, please describe how.*
	- No
15. *Does the dataset contain data that might be considered sensitive in any way (for example, data that reveals race or ethnic origins, sexual orientations, religious beliefs, political opinions or union memberships, or locations; financial or health data; biometric or genetic data; forms of government identification, such as social security numbers; criminal history)? If so, please provide a description.*
	- Not as a personal attribute
16. *Any other comments?*
	- None

**Collection process**

1. *How was the data associated with each instance acquired? Was the data directly observable (for example, raw text, movie ratings), reported by subjects (for example, survey responses), or indirectly inferred/derived from other data (for example, part-of-speech tags, model-based guesses for age or language)? If the data was reported by subjects or indirectly inferred/derived from other data, was the data validated/verified? If so, please describe how.*
	- Not specified
2. *What mechanisms or procedures were used to collect the data (for example, hardware apparatuses or sensors, manual human curation, software programs, software APIs)? How were these mechanisms or procedures validated?*
	- Not specified
3. *If the dataset is a sample from a larger set, what was the sampling strategy (for example, deterministic, probabilistic with specific sampling probabilities)?*
	- No specified
4. *Who was involved in the data collection process (for example, students, crowdworkers, contractors) and how were they compensated (for example, how much were crowdworkers paid)?*
	- Us population residents
5. *Over what timeframe was the data collected? Does this timeframe match the creation timeframe of the data associated with the instances (for example, recent crawl of old news articles)? If not, please describe the timeframe in which the data associated with the instances was created.*
	- 1950-2022, however, it only includes up to 2018 in the data. 
6. *Were any ethical review processes conducted (for example, by an institutional review board)? If so, please provide a description of these review processes, including the outcomes, as well as a link or other access point to any supporting documentation.*
	- Not specified
7. *Did you collect the data from the individuals in question directly, or obtain it via third parties or other sources (for example, websites)?*
	- Online website
8. *Were the individuals in question notified about the data collection? If so, please describe (or show with screenshots or other information) how notice was provided, and provide a link or other access point to, or otherwise reproduce, the exact language of the notification itself.*
	- None
9. *Did the individuals in question consent to the collection and use of their data? If so, please describe (or show with screenshots or other information) how consent was requested and provided, and provide a link or other access point to, or otherwise reproduce, the exact language to which the individuals consented.*
	- Not specified
10. *If consent was obtained, were the consenting individuals provided with a mechanism to revoke their consent in the future or for certain uses? If so, please provide a description, as well as a link or other access point to the mechanism (if appropriate).*
	- Data is available for all public to see
11. *Has an analysis of the potential impact of the dataset and its use on data subjects (for example, a data protection impact analysis) been conducted? If so, please provide a description of this analysis, including the outcomes, as well as a link or other access point to any supporting documentation.*
	- none
12. *Any other comments?*
	- None

**Preprocessing/cleaning/labeling**

1. *Was any preprocessing/cleaning/labeling of the data done (for example, discretization or bucketing, tokenization, part-of-speech tagging, SIFT feature extraction, removal of instances, processing of missing values)? If so, please provide a description. If not, you may skip the remaining questions in this section.*
	- Yea. There was removal of some columns that indicated sections of that rows that were categorized into one. renamed some of the column names for easier access and understanding. Filtered out many of the not needed information for the purpose of what data i wanted to create.
2. *Was the "raw" data saved in addition to the preprocessed/cleaned/labeled data (for example, to support unanticipated future uses)? If so, please provide a link or other access point to the "raw" data.*
	- Yes. It was aved as a csv file, and further downloading instructions can be found in scripts/02-download_data.R
3. *Is the software that was used to preprocess/clean/label the data available? If so, please provide a link or other access point.*
	- Yes. We used arrow to save data as parquet file
4. *Any other comments?*
	- None

**Uses**

1. *Has the dataset been used for any tasks already? If so, please provide a description.*
	- In the report: Exploring Suicide Rates: Assessing Age, Sex and Race Patterns with Statistical Insights into Suicide Risk. Can be found in paper/paper.pdf
2. *Is there a repository that links to any or all papers or systems that use the dataset? If so, please provide a link or other access point.*
	- https://github.com/sarahhhh02/suicides.git
3. *What (other) tasks could the dataset be used for?*
	- For analysing possible risk factors for suicide rates. Used for public health and government to help witht he increasing numbers. 
4. *Is there anything about the composition of the dataset or the way it was collected and preprocessed/cleaned/labeled that might impact future uses? For example, is there anything that a dataset consumer might need to know to avoid uses that could result in unfair treatment of individuals or groups (for example, stereotyping, quality of service issues) or other risks or harms (for example, legal risks, financial harms)? If so, please provide a description. Is there anything a dataset consumer could do to mitigate these risks or harms?*
	- None
5. *Are there tasks for which the dataset should not be used? If so, please provide a description.*
	- None
6. *Any other comments?*
	- None

**Distribution**

1. *Will the dataset be distributed to third parties outside of the entity (for example, company, institution, organization) on behalf of which the dataset was created? If so, please provide a description.*
	- No
2. *How will the dataset be distributed (for example, tarball on website, API, GitHub)? Does the dataset have a digital object identifier (DOI)?*
	- No
3. *When will the dataset be distributed?*
	- Never
4. *Will the dataset be distributed under a copyright or other intellectual property (IP) license, and/or under applicable terms of use (ToU)? If so, please describe this license and/ or ToU, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms or ToU, as well as any fees associated with these restrictions.*
	- no
5. *Have any third parties imposed IP-based or other restrictions on the data associated with the instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms, as well as any fees associated with these restrictions.*
	- No
6. *Do any export controls or other regulatory restrictions apply to the dataset or to individual instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any supporting documentation.*
	- No
7. *Any other comments?*
	- No

**Maintenance**

1. *Who will be supporting/hosting/maintaining the dataset?*
	- Sarah Lee
2. *How can the owner/curator/manager of the dataset be contacted (for example, email address)?*
	- sarahhhh.lee@mail.utoronto.ca
3. *Is there an erratum? If so, please provide a link or other access point.*
	- None
4. *Will the dataset be updated (for example, to correct labeling errors, add new instances, delete instances)? If so, please describe how often, by whom, and how updates will be communicated to dataset consumers (for example, mailing list, GitHub)?*
	- No
5. *If the dataset relates to people, are there applicable limits on the retention of the data associated with the instances (for example, were the individuals in question told that their data would be retained for a fixed period of time and then deleted)? If so, please describe these limits and explain how they will be enforced.*
	- None
6. *Will older versions of the dataset continue to be supported/hosted/maintained? If so, please describe how. If not, please describe how its obsolescence will be communicated to dataset consumers.*
	- No
7. *If others want to extend/augment/build on/contribute to the dataset, is there a mechanism for them to do so? If so, please provide a description. Will these contributions be validated/verified? If so, please describe how. If not, why not? Is there a process for communicating/distributing these contributions to dataset consumers? If so, please provide a description.*
	- Can download from https://github.com/sarahhhh02/suicides.git as a starting point and build from there
8. *Any other comments?*
	- None

\newpage

# References
